A user-extensible and adaptable parser architecture

### 2008-05-28 doc/logparser-acronyms.tex
Corrections after proofreading.

\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
    SMTP is the protocol which transfers mail from the sender to the
    recipient across the Internet.  %XXX A brief introduction to SMTP is
    %provided in \sectionref{SMTP background}.
}}

@@

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at \cite{Wikipedia-sql}.
}}

@@

\newacronym{regex}{Regular Expression}{description={
    Regular expressions are a method of matching patterns in text that are
    explained in detail in~\cite{Wikipedia-regex, perlre}.
}}

@@

\newacronym{ATN}{Augmented Transition Networks}{description={
    Augmented Transition Networks are a tool used in Computational
    Linguistics; they are originally described in~\cite{atns} and further
    in~\cite{nlpip}.
}}

@@

\newacronym{NLP}{Natural Language Processing}{description={
    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
    our understanding of the languages normally used by humans (e.g.\
    English, Spanish, Japanese), with the goal of writing programs which
    can understand human languages.  Potential applications include:
    programs which will automatically summarise a text; programs which
    can be taught facts, will understand questions and formulate answers to
    those questions.  It is in no way connected to Neuro-linguistic
    programming, which is generally found when searching for NLP.
}}

>>>>

\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
    SMTP, originally defined in \RFC{}~821~\cite{RFC821} and updated in
    \RFC{}~2821~\cite{RFC2821}, is the protocol which transfers mail from
    the sender to the recipient across the Internet.  It is a simple, human
    readable, plain text protocol, making it quite easy to test and debug
    problems with it.  A detailed description of \SMTP{} is beyond the
    scope of this document: introductory guides can be found
    at~\cite{smtp-intro-01, smtp-intro-02}, the definitive references
    are~\cite{RFC821, RFC2821}.
    %XXX A brief introduction to SMTP is
    %provided in \sectionref{SMTP background}.
}}

@@

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at \cite{Wikipedia-sql}; a good introduction can be found
    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
    sqlcourse.com}.
}}

@@

\newacronym{regex}{Regular Expression}{description={
    Regular expressions are a method of matching patterns against text that
    are explained in detail in~\cite{Wikipedia-regex, perlre}.
}}

@@

\newacronym{ATN}{Augmented Transition Networks}{description={
    Augmented Transition Networks, originally described in~\cite{atns} and
    further in~\cite{nlpip}, are a tool used in Computational Linguistics
    for creating grammars to parse or generate sentences,
}}

@@

\newacronym{NLP}{Natural Language Processing}{description={
    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
    our understanding of the languages normally used by humans (e.g.\
    English, Spanish, Japanese), with the goal of writing programs which
    can understand human languages.  It is in no way connected to
    Neuro-linguistic programming, which is generally found when searching
    for NLP.
}}

### 2008-05-28 papers/sgai-2008/paper.tex
Corrections after proofreading.

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{
    \begin{figure}[hbt!]
        \caption{#2}\label{#3}
        \includegraphics{#1}
    \end{figure}
}

@@

\abstract{%
    Some parsers need to be very precise and strict in what they parse, yet
    must allow users to easily adapt existing rules or add new rules to
    parse new inputs, without requiring the user to have an in-depth
    knowledge and understanding of the parser's internal workings.  This
    paper presents a novel parsing architecture which aims to make the
    process of parsing new inputs as simple as possible, enabling users to
    trivially add new rules or adapt existing rules (to parse variants of
    existing inputs) and relatively easily add new actions (to do something
    useful with a previously unknown class of input).  The architecture
    provides a framework which actions and rules fit into: this framework
    manages the parsing process, provides support functions for the
    actions, and automatically performs some optimisations which improve
    the speed of parsing.
}

@@

The architecture described herein was developed as part of a larger project to
improve anti-spam techniques by analysing the performance of the set of filters
currently in use, optimising the order and membership of the set based on that
analysis, and developing new measures to supplement the existing filters where
there is a deficiency.  The approach chosen was to analyse the log files
produced by the mail server, rather than modifying the mail server to generate
performance statistics; this approach improves the chances of other sites
testing and utilising the software.  The need arose for a parser capable of
dealing with the huge variety of log lines produced by Postfix~\cite{postfix},
the \MTA{} in use:

@@

    \item External resources utilised by Postfix (e.g.\ policy
        servers~\cite{policy-servers} or \RBL{} lookups) can also change
        their messages, without warning.

@@

Existing parsers were considered, but their parsing coverage was too low,
their parsing too inexact, and/or they did not extract sufficient data for
the purposes of this project.\footnote{A review of the existing parsers
considered for this project will be included in the thesis.} The effort
required to adapt and improve an existing parser was judged to be greater
than the effort to write one from scratch, as the techniques used by the
existing parsers severely limited their potential: some ignored the
majority of log lines, parsing specific log lines with great accuracy, but
in an inextensible fashion; others sloppily parsed the majority of log
lines, but were incapable of distinguishing between log lines of the same
type, e.g.\ rejections, and they extracted only a portion of the data
required for this project.

The solution developed is conceptually simple: provide a few generic functions,
each capable of dealing with an entire \textit{category\/} of log lines (e.g.\
rejections), accompanied by a multitude of precise patterns, each of which
matches all log lines of a specific type, and only that type (e.g.\ rejection
by a specific \RBLshort{}).

@@

    \item [Framework]  The framework is the structure which actions and
        rules fit into.  The framework provides the parsing loop, loading
        and validation of rules, shared data storage, storage of results,
        and other support functions.

@@

Decoupling also creates a clear separation of functionality: rules handle low
level details of identifying log lines and extracting data from a log line;
actions handle the higher level details of following the path a mail takes
through Postfix, assembling the required data, dealing with complications
arising, etc; the framework provides services to actions and stores data.

@@

The framework provides utility and support functions for the actions and
rules.  It provides shared storage to pass data between actions, most
notably the data structures used to represent each connection and mail,
though there are some others.  It loads and validates rules, controls
parsing of log files, invoking actions and tracking performance data, and
stores results in the database on request.  The framework is independent of
the parser being written: the same functionality will be required from the
framework for most or all parsers.

The function which finds the rule matching the input line and invokes the
requested action can be expressed in pseudo-code (with indentation denoting
flow of control) as:

@@

@@

Actions contain the code required to deal with log lines from a particular
category, e.g.\ rejections.  The actions are parser-specific: each parser
will need to write the actions it requires from scratch, unless it's
extending an existing parser.  In the Postfix parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules.
The distribution of rules to actions is very skewed, as can be seen in
Fig.~\refwithpage{Distribution of rules per action}.
\showgraph{build/plot-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} The distribution is skewed for
the same reason the architecture is so successful in enabling users to
parse new inputs: most new inputs can be parsed by a combination of a new
rule and an existing action.  Unsurprisingly the action with the highest
number of associated rules is \texttt{REJECTION}, the action which handles
rejections; it's followed by \texttt{INFO}, the action responsible for
handling informative inputs: these are log lines that supplement the data
gathered from other log lines, but that don't themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: there are many log lines which are not of interest to the
parser, e.g.\ warning messages about \DNS{} mismatches, or log lines
lacking a way to identify which data structure they should be associated
with.\footnote{Postfix allocates a queueid (explained in the
glossary,~\sectionref{Glossary}) to each mail; most log lines contain the
queueid of the associated mail, but some log lines, particularly with older
versions of Postfix, lack a queueid, preventing their association with a
particular data structure.}  The parser warns about every unparsed input,
to alert the user that they need to alter or extend their rules, hence the
\texttt{IGNORE} action: the input has been parsed, but it is of no benefit,
so it is ignored.  The remaining actions have only one or two associated
rules: in some cases this is because there will only ever be one type of
input for that action, e.g.\ all log lines showing that a remote client has
connected are matched by a single rule and handled by the \texttt{CONNECT}
action; in other cases the action is required to address a deficiency in
the log files, or a complication which arises during
parsing.\footnote{\label{complications footnote}The complications which
arose during development of the parser will be documented in detail in the
author's thesis.}

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour or resolve
challenges, with all the attendant risks of adversely affecting the
existing, working parser, a new, independent action can be written to
satisfy the new requirements.  In some cases the new action will require
the cooperation of other actions, e.g.\ to set or check a flag; this does
introduce the possibility introducing failure by modifying existing
actions, but modifications should be smaller and less frequent, and thus
failures should be less likely and easier to test for and diagnose.  The
parser can be written in an Object
Oriented~\cite{Wikipedia-object-orientation} style, allowing sub-classes to
extend or override actions in addition to adding new actions; because each
action is an independent function, the sub-class need only modify the
action it is interested in, rather than reproducing large chunks of
functionality.

@@

    \item [program] The Postfix program whose log lines the rule applies
        to.  This avoids needlessly trying rules which won't match the log
        line, or worse, might match unintentionally.  Rules whose program
        is \texttt{*} will be tried against any log lines which aren't
        parsed by program specific rules.

@@

    \item [result\_data] Sometimes rules need to supply a piece of data
        which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format is the same
        as for \texttt{result\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for use with small amounts of
        data rather than large amounts in any one rule, so dealing with
        escape sequences was deemed unnecessary.} is permitted on the right
        hand side of the assignment.

@@

The architecture can be compared to existing approaches to aid in fully
understanding it.

@@

    \item [Context-free rules:]  Rules are context free: they do not take
        into account inputs which have previously been seen, nor do they
        perform any kind of lookahead to future inputs.  In context-free
        grammar terms the parser rules could be described as:

        $\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

    \item [Partially context-sensitive actions:]  XXX IS THIS THE CORRECT
        TERMINOLOGY\@?  Actions can consult the results of previous actions
        when executing, providing some context sensitivity.  In some cases
        the lack of results from previous actions can be useful
        information, e.g.\ in the Postfix parser when a client disconnects
        if there have been no useful log lines for that client the data is
        discarded, rather than trying to store empty records in the
        database.

    \item [Matching is simple:]  The first matching rule wins: there is no
        backtracking to try alternate rules, no attempt made to pick a
        \textit{best\/} rule; the first rule which matches the line is the
        rule which parses it and determines the action that will be
        invoked.\footnote{Rule authors can affect the order rules will be
        tried in by setting the \texttt{priority} attribute of one or more
        rules.}

    \item [Transduction:]  The parser can be thought of as implementing
        transduction: it takes data in one form (log files) and transforms
        it to another form: a database in the case of the Postfix log
        parser, though it could be normalised files or files which only
        contain discrepancies or other interesting data.

    \item [Closer to \NLP{} than fixed grammar:]  Unlike traditional
        parsers such as those used when compiling a programming language,
        this architecture does not have a fixed grammar specification which
        inputs must adhere to.  The architecture is designed for easy user
        alteration, in fact such alteration is expected.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would have much larger log files on a daily basis).  When
generating the data for the graphs included in this document,
\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
log lines) were each parsed 10 times, the first run discarded, and the
execution time for the remaining 9 runs averaged.  The first run is
discarded for two reasons:

@@

    \item [shuffle] Random ordering --- the rules will be shuffled once
        before use and will retain that ordering for the entirety of the
        log file.  Note that the ordering will change every time the parser
        is executed, so 10 different orderings will be generated for each
        log file in the test run.  This is intended to represent an
        unsorted rule set.

@@

Figure~\refwithpage{percentage increase of shuffled and reversed over
normal} shows the percentage increase of execution times.  Overall this
provides a modest but worthwhile performance increase of over 10\%, for a
small investment in time and programming.
\showgraph{build/plot-normal-shuffle-reverse-factor}{Percentage increase of
shuffled and reversed over normal}{percentage increase of shuffled and
reversed over normal}

>>>>

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    \begin{figure}[hbt!]
        \caption{#2}\label{#3}
        \includegraphics{#1}
    \end{figure}
}

@@

\abstract{%
    Some parsers need to be very precise and strict in what they parse, yet
    must allow users to easily adapt existing rules or add new rules to
    parse new inputs, without requiring that user have an in-depth
    knowledge and understanding of the parser's internal workings.  This
    paper presents a novel parsing architecture which aims to make the
    process of parsing new inputs as simple as possible, enabling users to
    trivially add new rules or adapt existing rules (to parse variants of
    existing inputs) and relatively easily add new actions (to process a
    previously unknown class of input).  The architecture provides a
    framework which actions and rules fit into: this framework manages the
    parsing process, provides support functions for the actions, and
    automatically performs some optimisations which improve the speed of
    parsing.
}

@@

The architecture described herein was developed as part of a larger project
to improve anti-spam techniques by analysing the performance of the set of
filters currently in use, optimising the order and membership of the set
based on that analysis, and developing new measures to supplement the
existing filters where there is a deficiency.  The approach chosen to
measure performance was to analyse the log files produced by the mail
server, rather than modifying the mail server to generate performance
statistics; this approach improves the chances of other sites testing and
utilising the software.  The need arose for a parser capable of dealing
with the huge variety of log lines produced by Postfix~\cite{postfix}, the
mail server in use:

@@

    \item External resources utilised by Postfix (e.g.\ policy
        servers~\cite{policy-servers} or \RBL{} lookups) can also change
        their messages, sometimes without warning.

@@

Existing parsers were considered, but their parsing coverage was too low,
their parsing too inexact, and/or they did not extract sufficient data for
the purposes of this project.\footnote{A review of the existing parsers
considered for this project will be included in the author's thesis.} The
effort required to adapt and improve an existing parser was judged to be
greater than the effort to write one from scratch, as the techniques used
by the existing parsers severely limited their potential: some ignored the
majority of log lines, parsing specific log lines with great accuracy, but
in an inextensible fashion; others sloppily parsed the majority of log
lines, but were incapable of distinguishing between log lines of the same
type, e.g.\ rejections, and they extracted only a portion of the data
required for this project.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific
\RBLshort{}).

@@

    \item [Framework]  The framework is the structure which actions and
        rules fit into.  It provides the parsing loop, loading and
        validation of rules, shared data storage, storage of results, and
        other support functions.

@@

Decoupling also creates a clear separation of functionality: rules handle
low level details of identifying and extracting data from log lines;
actions handle the higher level details of following the path a mail takes
through Postfix, assembling the required data, dealing with complications
arising, etc; the framework provides services to actions and stores data.

@@

The framework provides utility and support functions for the actions and
rules.  It provides shared storage to pass data between actions, most
notably the data structures used to represent each connection and mail.
It loads and validates rules, controls
parsing of log files, invokes actions, tracks how often each rule matches
(to optimise rule ordering, see \sectionref{Rule ordering}), and
stores results in the database on request.  The framework is independent of
the parser being written: the same functionality will be required from the
framework for most or all parsers.

The function which finds the rule matching the log line and invokes the
requested action can be expressed in pseudo-code (with indentation denoting
flow of control) as:

@@

The framework takes care of miscellaneous support functions and low level
details of parsing, freeing the authors of actions to concentrate on
writing productive code.  It links actions and rules, allowing either to be
improved independently of the other.

@@

Actions contain the code required to deal with log lines from a particular
category, e.g.\ rejections.  The actions are parser-specific: each parser
will need to write the actions it requires from scratch, unless it's
extending an existing parser.  In the Postfix parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules;
the distribution of rules to actions is very skewed, as shown in
Fig.~\refwithpage{Distribution of rules per action}.
\showgraph{build/plot-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} The distribution is skewed for
the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  Unsurprisingly the action with the
highest number of associated rules is \texttt{REJECTION}, the action which
handles rejections; it's followed by \texttt{INFO}, the action responsible
for handling informative inputs: these are log lines that supplement the
data gathered from other log lines (such as rejections), but that don't by
themselves represent interesting events.  The third most common action is,
perhaps surprisingly, \texttt{IGNORE}: there are many log lines which are
not of interest to the parser, e.g.\ warning messages about DNS mismatches,
or log lines lacking a way to identify which data structure they should be
associated with.\footnote{Postfix allocates a queueid (explained in the
glossary,~\sectionref{Glossary}) to each mail; most log lines contain the
queueid of the associated mail, but some log lines (particularly with older
versions of Postfix) lack a queueid, preventing their association with a
particular data structure.}  The parser warns about every unparsed log
line, to alert the user that they need to alter or extend their rules; the
\texttt{IGNORE} action does nothing when executed, allowing uninteresting
log lines to be parsed without any effects.  The remaining actions have
only one or two associated rules: in some cases this is because there will
only ever be one type of input for that action, e.g.\ all log lines showing
that a remote client has connected are matched by a single rule and handled
by the \texttt{CONNECT} action; in other cases the action is required to
address a deficiency in the log files, or a complication which arises
during parsing.\footnote{\label{complications footnote}The complications
which arose during development of the parser will be documented in detail
in the author's thesis.}

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour or resolve
challenges, with all the attendant risks of adversely affecting the
existing, working parser, when a new requirement arises a new, independent
action can be written to satisfy it.  In some cases the new action will
require the cooperation of other actions, e.g.\ to set or check a flag;
there is a possibility of introducing failure by modifying existing
actions, but modifications will be smaller and occur less frequently than
with a monolithic architecture, and thus failures should be less likely and
easier to test for and diagnose.  The parser can be written in an Object
Oriented~\cite{Wikipedia-object-orientation} style, allowing sub-classes to
extend or override actions in addition to adding new actions; because each
action is an independent function, the sub-class need only modify the
action it is interested in, rather than reproducing large chunks of
functionality.

@@

    \item [program] The Postfix program whose log lines the rule applies
        to.  This avoids needlessly trying rules which won't match the log
        line, or worse, might match unintentionally.  Rules whose program
        is \texttt{*} will be tried against any log lines which aren't
        parsed by program specific rules, allowing generic rules to be
        written.

@@

    \item [result\_data] Sometimes rules need to supply a piece of data
        which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format is the same
        as for \texttt{result\_cols}, except that arbitrary data is
        permitted on the right hand side of the assignment.

@@

@@

    \item [Context-free rules:]  Rules are context free: they do not take
        into account inputs which have previously been seen, nor do they
        perform any kind of lookahead to future inputs.  In context-free
        grammar terms the parser rules could be described as:
        $\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

    \item [Partially context-sensitive actions:]  XXX IS THIS THE CORRECT
        TERMINOLOGY\@?  Actions can consult the results of previous actions
        when executing, providing some context sensitivity.  In some cases
        the lack of results from previous actions can be useful
        information, XXX IS THE EXAMPLE NECESSARY\@?  E.g.\ in the Postfix
        parser when a client disconnects if there have been no useful log
        lines for that client the data is discarded, rather than trying to
        store empty records in the database.

    \item [Matching rules to log lines is simple:]  The first matching rule
        wins: there is no backtracking to try alternate rules, no attempt
        is made to pick a \textit{best\/} rule; the first rule which
        matches the line is the rule which parses it and determines the
        action that will be invoked.

    \item [Transduction:]  The parser can be thought of as implementing
        transduction: it takes data in one form (log files) and transforms
        it to another form (a database).  Other implementations are free to
        choose the output format most suitable for their needs.

    \item [Closer to \NLP{} than fixed grammar:]  Unlike traditional
        parsers such as those used when compiling a programming language,
        this architecture does not have a fixed grammar specification which
        inputs must adhere to.  The architecture is designed for easy user
        alteration, in fact such alteration is expected.  XXX NEED MORE
        HERE\@.

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push
        unused input back onto the input stream.  This was not a deliberate
        decision, rather a consequence of the line oriented nature of the
        log files; other implementations could pursue a more flexible
        approach.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would produce much larger log files on a daily basis).  When
generating the data for the graphs included in this document,
\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
log lines) were each parsed 10 times, the first run discarded, and the
execution time for the remaining 9 runs averaged.  XXX DO I NEED TO EXPLAIN
WHY\@?  The first run is discarded for two reasons:

@@

    \item [shuffle] Random ordering, intended to represent an unsorted rule
        set.  The rules will be shuffled once before use and will retain
        that ordering for the entirety of the log file.  Note that the
        ordering will change every time the parser is executed, so 10
        different orderings will be generated for each log file in the test
        run.

@@

Figure~\refwithpage{percentage increase of shuffled and reversed over
normal} shows the percentage increase of execution times for shuffled and
reversed rules.  Overall this optimisation provides a modest but worthwhile
performance increase of over 10\%, for a small investment in time and
programming.
\showgraph{build/plot-normal-shuffle-reverse-factor}{Percentage increase of
execution time for different rule orderings}{percentage increase of
shuffled and reversed over normal}
